---
output: 
  github_document:
    df_print: kable
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```


<!-- README.md is generated from README.Rmd. Please edit that file -->

# cbsodata4

<!-- badges: start -->

[![CRAN
status](https://www.r-pkg.org/badges/version/cbsodata4)](https://cran.r-project.org/package=cbsodata4)
[![Travis build
status](https://travis-ci.org/statistiekcbs/cbsodata4.svg?branch=master)](https://travis-ci.org/statistiekcbs/cbsodata4)
[![AppVeyor build
status](https://ci.appveyor.com/api/projects/status/github/statistiekcbs/cbsodata4?branch=master&svg=true)](https://ci.appveyor.com/project/edwindj/cbsodata4)
[![Codecov test coverage](https://codecov.io/gh/statistiekcbs/cbsodata4/branch/master/graph/badge.svg)](https://codecov.io/gh/statistiekcbs/cbsodata4?branch=master)

<!-- badges: end -->

The goal of cbsodata4 is allowing access to the Open Data API v4 of Statistics Netherlands, which is 
in beta testing.

It is the successor of R package [cbsodataR](https://CRAN.R-project.org/package=cbsodataR).

- It uses the new / more efficient OData4 API

- The function names for `cbs4_add_*` are more consistent.

- The download storage is faster (using `data.table`)

- It offers a familiar wide data format for users (`cbs4_get_data`) as well as the default long format (`cbs4_get_observations`).

## Installation

<!--

You can install the released version of cbsodata4 from [CRAN](https://CRAN.R-project.org) with:

```r
install.packages("cbsodata4")
```
!-->

At the moment only a development version can be installed from
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("statistiekcbs/cbsodata4")
```

## Usage

```{r}
library(cbsodata4)
```

Retrieve a list of all CBS / Statistics Netherlands tables with `cbs4_get_datasets`:

```{r}
# download the set of datasets
datasets <- cbs4_get_datasets()
datasets[1:5, c("Identifier", "Title")]
```

Get metadata of table `<Identifier>` (e.g. 60006) with `cbs4_get_metadata`:

```{r}
meta <- cbs4_get_metadata(id="60006")
print(meta)
meta$Properties$Title
# topics / measures
meta$MeasureCodes[,c("Identifier", "Title")]
# dimensions
meta$Dimensions[, c("Identifier", "Title")]
```

Retrieve data with `cbs4_get_data`:

```{r}
# wide format, each measure its own column
data <- cbs4_get_data("60006")
head(data[, 1:4])
```

or `cbs4_get_observations`

```{r}
# long format, one Value column
obs <- cbs4_get_observations("60006")
head(obs)
```

Add labels to a dataset with `cbs4_add_label_columns`:

```{r}
data <- cbs4_add_label_columns(data)
head(data[,1:5])

obs <- cbs4_add_label_columns(obs)
head(obs)
```


Find non-standard CBS catalogs with `cbs4_get_catalogs`:

```{r}
catalogs <- cbs4_get_catalogs()
catalogs[,c("Identifier", "Title")]
```

For more information see `vignette("cbsodata4")`
